**Title:** Exploring the effects of BCG vaccination in patients diagnosed with tuberculosis: observational study using the Enhanced Tuberculosis Surveillance system

### 2018-06-13 output/Paper.Rmd
added additional references and discussion of meta-analyses

**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (250), **Paper** 2740 (3500)

@@

**Keywords:** Tuberculosis, BCG, Surveillance

**What is the key question?**

Is there evidence in the Enhanced Tuberculosis (TB) Surveillance System that BCG vaccination improves TB outcomes?

**What is the bottom line?**

This analysis suggests that vaccination with BCG is associated with reduced all-cause mortality for TB patients and potentially a reduced chance of having a repeat episode.

**Why read on?**

BCG vaccination could reduce all-cause mortality in TB cases and may decrease the number of repeat episodes in low incidence countries.

@@

BCG’s primary mode of action is to directly prevent the development of active, symptomatic disease. Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[@Mangtani2014a] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit. Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG vaccinate at birth.[@Zwerling2011] Adult vaccination is no longer common in the UK, where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.

Vaccination policy has been primarily based on reducing the incidence of active TB and little attention has been given to any additional effects of BCG.[@Fine2005a; @Teo2006] There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[@Kleinnijenhuis2012] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[@Jeremiah2010] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[@Higgins] and both reduced TB[@Abubakar2013] and all-cause[@Rieckmann2016] mortality in the general population. Given that the immunology behind TB immunity is not well understood these findings suggest that BCG may play a more important role in improving TB outcomes than previously thought. We aimed to quantify the effects of BCG vaccination on outcomes for individuals with notified TB in England using routinely collected surveillance data to provide evidence for appropriate public health action and provision. Where we found an association, we additionally explored the role of years since vaccination, and age at vaccination.

@@

Little work has been done to assess the overall effect of BCG on outcomes for active TB cases although the possible non-specific effects of BCG are an area of active research.[@Higgins; @Kandasamy2016; @Pollard2017] Whilst multiple studies have investigated BCG's association with all-cause mortality, it has been difficult to assess whether the association continues beyond the first year of life.[@Pollard2017] The effect size of the association we identified between BCG and all-cause mortality in active TB cases was comparable to that found in a Danish case-cohort study in the general population (aHR: 0.58 (95% CI 0.39 to 0.85).[@Rieckmann2016] A recent systematic review also found that BCG vaccination was associated with reduced all-cause mortality in neonates, with an average relative risk of 0.70 (95% CI 0.49 to 1.01) from five clinical trials and 0.47 (95% CI 0.32 to 0.69) from nine observational studies at high risk of bias.[@Higgins] We found some weak evidence that BCG vaccination was associated with reduced deaths due to TB (in those who died), although our point estimate had large confidence intervals. Several meta-analyses have found evidence supporting this association,[@Colditz1994; @Abubakar2013] with one meta-analysis estimating a 71% (RR: 0.29 95% CI 0.16 to 0.53) reduction in deaths due to TB.[@Colditz1994] Our study was limited by the relative rarity of deaths due to TB in England, whereas other approaches benefited from the use of meta-analysis. This study could not determine the possible causal pathway for the association between BCG vaccination all-cause mortality, and recurrent TB. These are important to establish in order to understand the effect of BCG vaccination on TB outcomes.

>>>>

**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (300), **Paper** 2740 (5000)

@@

**Keywords:** Tuberculosis, BCG, Surveillance, Non-specific, Mortality

**Highlights**

- Is there evidence in the Enhanced Tuberculosis (TB) Surveillance System that BCG vaccination improves TB outcomes?

- This analysis suggests that vaccination with BCG is associated with reduced all-cause mortality for TB patients and potentially a reduced chance of having a repeat episode. There was little evidence of an association with other TB outcomes.

- BCG vaccination could reduce all-cause mortality in TB cases and may decrease the number of repeat episodes in low incidence countries.

@@

BCG’s primary mode of action is to directly prevent the development of active, symptomatic disease. Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[@Mangtani2014a] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[@Barreto2014a] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG vaccinate at birth.[@Zwerling2011; @WHOBCG2018] Adult vaccination is no longer common in the UK, where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.

Vaccination policy has been primarily based on reducing the incidence of active TB and little attention has been given to any additional effects of BCG.[@Fine2005a; @Teo2006] There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[@Kleinnijenhuis2012] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[@Jeremiah2010] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[@Garly2003; @Higgins] and both reduced TB[@Abubakar2013] and all-cause[@Rieckmann2016] mortality in the general population. Given that the immunology behind TB immunity is not well understood these findings suggest that BCG may play a more important role in improving TB outcomes than previously thought. We aimed to quantify the effects of BCG vaccination on outcomes for individuals with notified TB in England using routinely collected surveillance data to provide evidence for appropriate public health action and provision. Where we found an association, we additionally explored the role of years since vaccination, and age at vaccination.

@@

Little work has been done to assess the overall effect of BCG on outcomes for active TB cases although the possible non-specific effects of BCG are an area of active research.[@Higgins; @Kandasamy2016; @Pollard2017] Whilst multiple studies have investigated BCG's association with all-cause mortality, it has been difficult to assess whether the association continues beyond the first year of life.[@Pollard2017] The effect size of the association we identified between BCG and all-cause mortality in active TB cases was comparable to that found in a Danish case-cohort study in the general population (aHR: 0.58 (95% CI 0.39 to 0.85).[@Rieckmann2016] A recent systematic review also found that BCG vaccination was associated with reduced all-cause mortality in neonates, with an average relative risk of 0.70 (95% CI 0.49 to 1.01) from five clinical trials and 0.47 (95% CI 0.32 to 0.69) from nine observational studies at high risk of bias.[@Higgins] We found some weak evidence that BCG vaccination was associated with reduced deaths due to TB (in those who died), although our point estimate had large confidence intervals. Several meta-analyses have found evidence supporting this association,[@Colditz1994; @Abubakar2013] with one meta-analysis estimating a 71% (RR: 0.29 95% CI 0.16 to 0.53) reduction in deaths due to TB in individuals vaccinated with BCG.[@Colditz1994] The meta-analysis performed by Abubakar et al. also found consistent evidence for this association but was limited by the number, and quality, of studies available.[@Abubakar2013] In contrast to our study, both of these meta-analyses estimated the protection from TB mortality in BCG vaccinated individuals rather than in BCG vaccinated cases. Additionally, neither study explored the association between BCG vaccination and all-cause mortality or recurrent TB. This study could not determine the possible causal pathway for the association between BCG vaccination all-cause mortality, and recurrent TB. These are important to establish in order to understand the effect of BCG vaccination on TB outcomes.
